```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install additional packages

Packages extent the functionality of R. Currently, there are more the 7000 packages available on the main repository [CRAN](https://cran.r-project.org/).

To install a package simply type

```{r install_vegan, eval=FALSE}
install.packages('vegan')
```

into the console and hit enter.
This will download and **install** the `vegan` package to your computer and you should see something like this:

```{r install_vegan2, eval=FALSE}
Installing package into ‘/home/edisz/R_libs/all’
(as ‘lib’ is unspecified)
trying URL 'https://cran.rstudio.com/src/contrib/vegan_2.3-0.tar.gz'
Content type 'application/x-gzip' length 2059624 bytes (2.0 MB)
==================================================
downloaded 2.0 MB

* installing *source* package ‘vegan’ ...
** package ‘vegan’ successfully unpacked and MD5 sums checked
** libs
[...]
** testing if installed package can be loaded
* DONE (vegan)
```

After installing the package, you need to **load** the package to use its functionality, which is done with the `library()` function.

```{r load_vegan, message=TRUE}
library('vegan')
```


