```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## First steps in R

R is a powerful calculator with standard arithmetic operators:

```{r}
# Addition
1 + 6
```

```{r}
# Multiplication
2 * 6
```

```{r}
# multiplication and division first, then addition and subtraction
2 * 6 + 1
```

```{r}
# use of brackets
2 * (6 + 1)
```

```{r}
# square root
sqrt(16)
```

```{r}
# powers
16^0.5
2^2
```

```{r}
# natural log
log(10)
```

```{r}
# decadic log
log10(10)
```

```{r}
# log to base of 4
log(10, 4)
```

```{r}
# exponential
exp(2)
```

```{r error=TRUE}
# pi is a built-in constant
pi
# but not e
e^2
```








If we want to refer to a result later on, we can give them a name.

```{r}
a <- 1 + 6
```

Here I *assign* (`<-`) the result of `1 + 6` to an *object* named `a`.
If you just type the name into the console R shows whats inside this object:

```{r}
a
```

You can also do calculations with this object:

```{r}
2 * a
```

Note, that R will overwrite objects:
```{r}
a
a <- 2*a
a
```

If R doesn't know about an object you get the following error:
```{r error=TRUE}
b
```

which says: *"I looked for the object 'b', but could not find it"*".

You can see objects that are available in the `Environment` pane of RStudio. Alternatively you can use the `ls()` function.


