```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Reading data into R


### Reading text files

One of the easiest ways to store data is to use text files (or comma separated files `.csv`). 
Nearly every tool can handle such data.

To read such data into R we use the function `read.table`.

`read.table` has a lot of arguments to control it's fumctionality:

```{r eval = FALSE}

read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = default.stringsAsFactors(),
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
```



The most useful ones are:

* `file` : character, the path to the file (either absolute or relative to the workign directory [`#! need link here`])
* `header` : logical (either `TRUE` or `FALSE`). Does the file include column names in the first row?
* `sep` : columns can be separated in text files by different characters (e.g. `,` or `;` or `|`)
* `dec` : the decimal separator. Default is `.`, but in some countries (like germany) this can also be `,`
* `stringsAsFactors` : Should be strings interpreted as factors? It tend to use `stringsAsFactors = FALSE` (default is `TRUE`) and convert only if needed.

So a typical command to load data would be:

```{r read_csv}
setwd( "/home/edisz/Documents/Uni/Projects/qetxrbook")
df <- read.table('data/iris1.csv', sep = ';', header = TRUE, stringsAsFactors = FALSE)
str(df)
```


### Writing text files

Writing text files is done via `write.table`. It has a similar set of attributes as above. 
However, to others should be mentioned:

* `append` : If `TRUE` the data is appended to the file. Otherwise, the existing file is overwritten.
* `row.names` : Is always use `row.names=FALSE`, because rownames should not contain information and the columns are better aligned.

So to write the above data.frame to a file use:

```{r eval=FALSE}
write.table(df, "/home/edisz/mytable.csv", sep = ';', row.names = FALSE)
```



### Excel files

Eventually you will have to work with Microsoft Excel files.
Excel file are [not a good format](http://www.win-vector.com/blog/2012/12/please-stop-using-Excel-like-formats-to-exchange-data/) to store data and have been hard to read into R.

Luckily, there is now the [`readxl`](https://cran.r-project.org/web/packages/readxl/index.html) package available on CRAN to make this easier. 




### Other files

R can also read a lof of other formats (SPSS, GIS-data, ...), however you'll certainly need additional packages. Just [search](../rintro/help.html) for it on the internet.



### Databases



### Exercises





